\documentclass[12pt,oneside]{article}
% \documentclass[12pt,twocolumn]{article}

\usepackage[utf8]{inputenc}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{bookmark}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{mdframed}
\usepackage{setspace}
\usepackage{pgfplots}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{minted}     % For syntax highlighting

\usepackage{geometry}
\geometry{
  a4paper,            % Paper size
  total={7.5in,10in}, % Width and height of the text area
  left=0.75in,        % Left margin
  right=0.75in,       % Right margin
  top=1in,            % Top margin
  bottom=1in          % Bottom margin
}

% \usepackage[absolute]{textpos}\TPGrid{16}{16}
\usepackage{tikz}
  \usetikzlibrary{shapes}
  \usetikzlibrary{arrows.meta,arrows,shadows,trees,fit,calc,positioning,decorations.pathmorphing}
\usepackage{xcolor}
  \definecolor{mypink}{HTML}{560CCE}
  \definecolor{myblack}{HTML}{232527}

  \definecolor{add}{rgb}{0,0.6,0}   % Green for additions
  \definecolor{remove}{rgb}{0.6,0,0} % Red for deletions

  \definecolor{graybg}{rgb}{0.95,0.95,0.95}

\usepackage{hyperref}
  \hypersetup{colorlinks=true,allcolors=blue!40!black}

\setlength{\topskip}{6pt}
\setlength{\parindent}{0pt} 
\setlength{\parskip}{6pt}
\setlength{\columnsep}{1cm} % Space between columns

\date{\small\today}
\title{%
  Optimizing Development Efficiency: The Case for AI-Powered Documentation \\
  \colorbox{mypink}{\small\sffamily\color{white}{White Paper}}}

\usepackage[style=authoryear,sorting=nyt,backend=biber,
  hyperref=true,abbreviate=true,
  maxcitenames=1,maxbibnames=1]{biblatex}

\renewbibmacro{in:}{}
\addbibresource{books.bib}

\author{Bohdan Snisar}

\begin{document}
\raggedbottom
\maketitle

%---------------------------
%         Abstract
%---------------------------
\begin{abstract}
Developers face significant challenges in understanding evolving codebases, worsened by fragmented workflows and rapid development. AI-assisted code generation adds complexity, increasing code churn and technical debt. This white paper introduces automated documentation powered by Large Language Models (LLMs) to address these issues. This approach ensures accuracy, maintainability, and alignment with business goals by generating high-fidelity initial documentation and incrementally updating it with each code change. Integrated into the Software Development Life Cycle (SDLC), automated documentation reduces cognitive load, enhances collaboration, and improves developer productivity, fostering clearer, more maintainable codebases.
\end{abstract}

%---------------------------
%    1. Introduction
%---------------------------
\section{Introduction}

The Software Development Life Cycle (SDLC) encompasses various activities— from requirements analysis and design to 
testing, deployment, and maintenance. These tasks are supported by a multitude of tools, such as issue trackers, code 
review platforms, and continuous integration (CI) systems, which often fragment developers’ workflows. As engineers 
switch between different tools and tasks, they face considerable difficulty forming a cohesive understanding of the 
product’s evolving codebase.

Despite a widespread perception that software engineers primarily focus on writing new code, research consistently 
shows a different reality: the bulk of a developer’s time is devoted to understanding existing code. For instance, 
up to 70\% of a developer’s effort centers on code comprehension, a process that frequently involves navigating 
numerous files \textcite{ko2006exploratory} and \textcite{Schroer_2024}. Internal studies at Microsoft confirm this 
trend, revealing that new code creation constitutes only a minor portion of developers’ daily work 
\textcite{meyer2021today}. This imbalance is further strained by the disconnect between high-level business objectives 
and low-level technical structures, complicating alignment and communication among cross-functional teams.

Recently, AI-assisted code generation has brought both opportunities and challenges to the forefront. On one hand, 
it promises enhanced productivity and faster development cycles; on the other, it leads to increased code churn and 
technical debt \textcite{pandi2023}. Rapid fixes, code duplication, and sparse documentation undermine maintainability. 
As \textcite{gitclear2024} indicates, AI-generated contributions often mirror the work habits of transient developers, 
fueling fragmented codebases and complicating efforts to align technical progress with broader strategic goals.

Reducing the cognitive overhead of understanding these complex, evolving codebases thus emerges as a critical 
necessity. Improved visualization tools and integrated developer tooling can mitigate some of these challenges, 
ensuring that engineers have seamless access to relevant information \textcite{meyer2021today}. In particular, 
automated documentation, powered by advanced AI models, can offer a transformative approach to code comprehension: 
delivering up-to-date, robust documentation that bridges the gap between technical minutiae and strategic intent.

\subsection{The Challenge of Manual Documentation}

Documentation serves as the foundation of knowledge sharing in software development, yet it 
presents persistent challenges in modern development environments:

\begin{itemize}
    \item Manual documentation creation diverts significant engineering resources from core development tasks, including feature implementation, bug resolution, and performance optimization.
    \item Sporadic documentation updates lead to accumulating inaccuracies and omissions, particularly in rapidly evolving codebases and those utilizing AI-generated code.
    \item The proliferation of documentation tools—including wikis, issue trackers, and code review systems—creates friction that often results in outdated or incomplete documentation.
    \item The increasing pace of development, coupled with AI-generated code contributions, amplifies documentation maintenance challenges and introduces inconsistent patterns.
\end{itemize}

As the pace of development accelerates, these problems become more pronounced, especially when developers are also 
contending with AI-generated code that may introduce inconsistent patterns or hidden complexities.

\subsection{Why Automated Documentation?}

Improving the \textbf{visibility} and \textbf{maintainability} of software projects hinges on reducing developers’ 
cognitive load and ensuring that essential information is easy to access and always up to date. 

\textbf{Automated documentation}, particularly when paired with Large Language Models (LLMs), offers a compelling 
solution:

\begin{itemize}
    \item \textbf{Context-Aware Generation:} LLMs can analyze codebases, and version histories, and commit messages to produce documentation that matches both technical realities and business goals.
    \item \textbf{Incremental Updates:} By monitoring repository changes, automated systems can generate patch-like documentation updates, eliminating the need for large-scale rewrites.
    \item \textbf{Transparency \& Alignment:} Near real-time updates keep cross-functional teams—ranging from product managers to QA—on the same page, ensuring that shifting business objectives are reflected in the code’s evolving architecture.
    \item \textbf{Reduced Technical Debt:} Documenting features and architectural decisions as they are introduced curbs the accumulation of stale or missing information—a significant contributor to technical debt \textcite{pandi2023}.
\end{itemize}

In the following sections, this white paper presents a structured methodology for LLM-powered automated documentation, 
detailing how it tackles the friction points of traditional manual approaches and how it can be woven into existing 
SDLC practices to maximize developer productivity and organizational agility.

%---------------------------
%    2. LLM-Powered Documentation: Overview
%---------------------------
\section{LLM-Powered Documentation: Overview}

\subsection{Initial Creation vs. Incremental Updates}
The central premise behind this solution is a two-phase approach:

\begin{enumerate}
  \item \textbf{Initial Documentation Generation}
  \begin{itemize}
    \item \emph{Objective}: Provide comprehensive coverage of the existing codebase.
    \item \emph{Outputs}: Class descriptions, method annotations, architectural overviews, and usage examples.
  \end{itemize}

  \item \textbf{Incremental Updates via Revision Patches}
  \begin{itemize}
    \item \emph{Objective}: Reflect changes in the code without rewriting entire documents.
    \item \emph{Outputs}: Small, focused documentation patches aligned with code diffs.
  \end{itemize}
\end{enumerate}

\subsection{Core Benefits}
\begin{enumerate}
  \item \textbf{Efficiency}: Minimizes manual labor by automating documentation tasks.
  \item \textbf{Accuracy}: Reduces the risk of human error by basing documentation on actual code diffs.
  \item \textbf{Relevance}: Ensures that documentation always mirrors the latest state of the codebase.
  \item \textbf{Scalability}: Handles large, complex codebases with minimal overhead.
  \item \textbf{Traceability}: Revision patches offer a clear historical record, enhancing collaboration and audits.
\end{enumerate}

%---------------------------
%         3. Methodology
%---------------------------
\section{Methodology}

\subsection{Step 1: Initial Documentation Generation}
\begin{enumerate}
  \item \textbf{Source Code Analysis}: The LLM inspects class structures, functions, and existing comments.
  \item \textbf{Context Gathering}: It synthesizes architectural overviews, method annotations, and usage examples.
  \item \textbf{Output Creation}: The system writes documentation describing input parameters, return types, default values, and exception handling scenarios.
\end{enumerate}

\subsection{Step 2: Tracking Changes and Generating Revision Patches}
\begin{enumerate}
  \item \textbf{Version Control Integration}: Monitors commits in Git or similar systems to identify differences between previous and current code versions.
  \item \textbf{Change Mapping}: Each code edit is linked to relevant sections in the existing documentation.
  \item \textbf{Patch Generation}: A concise “patch” is produced, updating only the sections affected by the code change.
\end{enumerate}

\subsubsection{JavaScript Example: Adding a New Parameter}
Below is a simplified JavaScript function before and after a new parameter (discount) is introduced. 
Following the code snippets, you’ll see a separate text block demonstrating how business context and technical 
details can be integrated into automatically generated documentation.

\noindent\textbf{Original Code:}
\begin{minted}[
  frame=lines,         % Adds an outline (top and bottom lines)
  bgcolor=graybg,      % Sets the background color
  linenos,             % Adds line numbers
  fontsize=\small,     % Adjusts font size
  breaklines           % Breaks long lines if needed
]{diff}
function calculateTotal(price, quantity) {
    return price * quantity;
}
\end{minted}

\noindent\textbf{Updated Code:}
The updated function introduces a new optional parameter, \texttt{discount}, and adjusts the total price accordingly.
Below is the Pull Request-style diff:

\begin{minted}[
  frame=lines,
  bgcolor=graybg,
  linenos,
  fontsize=\small,
  style=trac,
]{diff}
- function calculateTotal(price, quantity) {
+ function calculateTotal(price, quantity, discount = 0) {
     const total = price * quantity;
-    return total;
+    return total - (total * discount);
}
\end{minted}

Below is a before-and-after illustration of an automatically generated documentation block, followed by a \emph{diff} 
snippet resembling a Pull Request view.

\paragraph{Original Documentation Block (Before)}
\begin{verbatim}
  Function Name: calculateTotal
  
  Business Context:
  This function is a key part of the Invoicing module, which processes customer orders.
  By accurately computing final charges, it ensures proper billing and helps our Finance
  team maintain consistent revenue tracking.
  
  Parameters:
  1. price (Number)    - Represents the per-item cost.
  2. quantity (Number) - Number of items being purchased by the customer.
  
  Behavior:
  - Multiplies price by quantity to determine total.
  - Returns the final adjusted price for billing.
  
  Usage Example:
  calculateTotal(100, 2) -> returns 200
  
  Impact:
  - Accurate documentation ensures Sales and Finance can rely on correct billing data.
  - Potential risk if parameters are misconfigured or not applied properly.
  
  Revision Patch Overview:
  - (None so far; initial version)
\end{verbatim}


\paragraph{Modified Documentation Block (After)}
\begin{minted}[
    frame=lines,
    bgcolor=graybg,
    linenos,
    escapeinside=||,
    style=trac,
    numbersep=5pt,
    showspaces=false,
    tabsize=4,
    breaklines,
    breakanywhere,
    fontsize=\footnotesize,
    highlightcolor=add
]{diff}
 Function Name: calculateTotal

 Business Context:
 This function is a key part of the Invoicing module, which processes customer orders.
 By accurately computing final charges, it ensures proper billing and helps our Finance
-team maintain consistent revenue tracking.
+team maintain consistent revenue tracking. The optional discount parameter is commonly
+used for promotional events, loyalty programs, or corporate partner deals.

 Parameters:
 1. price (Number)    - Represents the per-item cost.
 2. quantity (Number) - Number of items being purchased by the customer.
+3. discount (Number) - An optional fraction (0–1) applied to the total purchase;
+                      default value is 0, indicating no discount.

 Behavior:
- - Multiplies price by quantity to determine total.
- - Returns the final adjusted price for billing.
+ - Multiplies price by quantity to determine total.
+ - Applies discount as a fraction of total if provided.
+ - Returns the final adjusted price for billing.

 Usage Example:
-calculateTotal(100, 2) -> returns 200
+1. Using no discount:
+   calculateTotal(100, 2) -> returns 200
+
+2. With a 10% discount:
+   calculateTotal(100, 2, 0.10) -> returns 180

 Impact:
 - Accurate documentation ensures Sales and Finance can rely on correct billing data.
-- Potential risk if parameters are misconfigured or not applied properly.
+- Potential risk if discount is misconfigured or not applied properly.

 Revision Patch Overview:
-- (None so far; initial version)
+- Added discount parameter to the function signature.
+- Adjusted documentation to reflect optional discount logic.
+- Updated usage examples to demonstrate discounted pricing.
\end{minted}

\subsection{Step 3: Seamless Integration}
\begin{itemize}
  \item \textbf{IDE Support}: Developers receive prompts for documentation patches directly in their IDE (e.g., VS Code, JetBrains).
  \item \textbf{Pull Requests}: Updated documentation patches appear in pull requests, ensuring peer review and alignment with newly introduced code.
  \item \textbf{CI/CD Pipelines}: Automated checks validate that the documentation is synchronized before deployments.
\end{itemize}

\subsection{Human-in-the-Loop and LLM Accuracy}
One of the most challenging aspects of using Large Language Models for documentation is ensuring 
the accuracy of the content they generate. While LLMs are powerful at synthesizing information, 
they are also prone to ``hallucinations,'' where confident-sounding but incorrect information may appear 
in the generated text. Moreover, quantifying the \emph{accuracy} of documentation relative to the actual 
codebase is non-trivial.

\paragraph{Human Review}
To mitigate these risks, this methodology integrates a human-in-the-loop review process:
\begin{enumerate}
  \item \textbf{Stakeholder Input}: Developers, product managers, or architects review generated documentation patches, validating correctness and business alignment.
  \item \textbf{Feedback Loop}: Approved changes feed back into the LLM’s context or fine-tuning dataset, improving future output.
  \item \textbf{Error Correction}: Any discovered inaccuracies are corrected manually and included in version-controlled documentation, ensuring a clear audit trail.
\end{enumerate}

\paragraph{Balancing Automation and Oversight}
While automation accelerates documentation upkeep, human oversight is crucial for validating business requirements 
against technical changes, catching edge cases or ambiguous code structures and Continuously refining the quality of 
LLM output and preventing drift.

By combining version-controlled documentation with a human-in-the-loop review framework, the approach achieves both \emph{speed} and \emph{reliability}, aligning with the dynamic nature of modern software development.



%---------------------------
% 4. Modular Documentation Through Slices
%---------------------------

\section{Modular Documentation Through Slices}

Modern software development necessitates not only robust tooling but also adaptive documentation strategies. This section introduces the \textbf{slice-based approach to documentation}, a modular and scalable framework designed to address the challenges of maintaining relevance, accessibility, and alignment across diverse stakeholder needs. By breaking down documentation into focused, interlinked units called \textit{slices}, this approach ensures precision and clarity in conveying both business and technical information.

\subsection{Slice: The Fundamental Unit of Documentation}

A \textbf{slice} is a self-contained, modular piece of documentation tied to a specific aspect of the codebase, such as a feature, workflow, or infrastructure component. Each slice serves as a \textit{Lego block} that, when combined with others, forms the complete documentation structure. The slice-based design addresses several critical needs:
\begin{itemize}
    \item \textbf{Granularity and Clarity:} Each slice focuses on a single functionality or process, ensuring clear, concise, and actionable documentation.
    \item \textbf{Stakeholder Accessibility:} Slices are designed to be understandable by both technical and non-technical stakeholders, bridging the gap between developers, product managers, and business teams.
    \item \textbf{Dynamic Evolution:} Slices update automatically with code changes, ensuring that the documentation remains accurate and reflective of the current system state.
\end{itemize}

Each slice is structured to offer a balanced mix of \textbf{business requirements documentation (BRD)} 
and \textbf{software decision record (SDR)} insights. Below is the standard structure:

\begin{enumerate}
    \item \textbf{Overview:} High-level purpose, business rationale, and technical summary.
    \item \textbf{Requirements and Constraints:} Business rules and technical limitations.
    \item \textbf{Workflow and Processes:} Step-by-step explanation, with business and technical perspectives.
    \item \textbf{Business and Technical Context:} Broader strategic and architectural relevance.
    \item \textbf{Contracts and Dependencies:} Input/output specifications, external dependencies.
    \item \textbf{State Management:} System-wide impacts of state changes and data interactions.
    \item \textbf{Code Mapping and Implementation:} Links to code artifacts and technical decisions.
\end{enumerate}

\subsection{Principles of Slice-Based Documentation}

The slice-based approach adheres to the following principles:
\begin{itemize}
    \item \textbf{Focused:} Each slice documents a single functionality, ensuring precision.
    \item \textbf{Versioned and Timestamped:} Slices evolve with the codebase, tracking changes with every commit.
    \item \textbf{Stakeholder-Friendly:} Designed for both technical (e.g., developers) and non-technical (e.g., product managers) audiences.
    \item \textbf{Rich Context:} Incorporates metadata and domain-specific insights.
    \item \textbf{Human-in-the-Loop Workflow:} A structured review process ensures accuracy and builds trust.
\end{itemize}

\subsection{Benefits of Slice-Based Documentation}

\begin{enumerate}
    \item \textbf{Improved Collaboration:} By providing contextually rich, accessible documentation, slices enhance communication across teams.
    \item \textbf{Reduced Cognitive Load:} Developers and stakeholders can focus on relevant slices, minimizing the effort required to understand the codebase.
    \item \textbf{Scalability:} The modular nature of slices supports large, evolving codebases without overwhelming documentation systems.
    \item \textbf{Traceability:} Each slice is tied to specific commits, workflows, and dependencies, providing a clear audit trail.
\end{enumerate}

\subsection{Integration with LLM-Powered Automation}

Slices form the foundation of the proposed LLM-powered automated documentation system. By leveraging LLMs to generate, update, and refine slices, the framework achieves:
\begin{itemize}
    \item \textbf{Initial Slice Creation:} Comprehensive documentation of existing functionality, workflows, and infrastructure.
    \item \textbf{Incremental Updates:} Patch-like modifications that reflect code changes.
    \item \textbf{Enhanced Accuracy:} Context-aware LLM capabilities ensure alignment with both technical realities and business goals.
\end{itemize}

\subsection{Example: Slice for a Code Logic Component}

\noindent\textbf{Component:} \texttt{calculateTotal} function in JavaScript.

\begin{itemize}
    \item \textbf{Overview:}
        \begin{itemize}
            \item \textbf{Purpose:} Calculates the total price for a transaction, including an optional discount.
            \item \textbf{Business Need:} Ensures accurate billing for customer orders.
            \item \textbf{Technical Summary:} Implements pricing logic, with discount support for promotions.
        \end{itemize}
    \item \textbf{Workflow and Processes:}
        \begin{itemize}
            \item \textbf{Business Process:} Applies pricing rules, including discount policies.
            \item \textbf{Technical Process:} Multiplies price and quantity, adjusts for discounts.
        \end{itemize}
    \item \textbf{State Management:}
        \begin{itemize}
            \item \textbf{System Impact:} Updates financial transaction records with calculated totals.
            \item \textbf{Data Relationships:} Links to order details and promotion tables.
        \end{itemize}
    \item \textbf{Code Mapping:}
        \begin{itemize}
            \item \textbf{Files:} \texttt{pricing.js}
            \item \textbf{Function:} \texttt{calculateTotal(price, quantity, discount)}
        \end{itemize}
\end{itemize}

This modular, slice-based framework elevates documentation from a static artifact to a dynamic, integral
component of the development lifecycle, ensuring that it remains relevant, actionable, and aligned with 
both technical and business needs.

%---------------------------
%         5. Conclusion
%---------------------------

\section{Conclusion}
Automated documentation powered by Large Language Models (LLMs) offers a viable path to reducing the cognitive overhead 
associated with modern software development. 
By continuously generating and updating high-fidelity documentation, organizations can ensure that their codebases 
remain transparent, maintainable, and aligned with ever-evolving business objectives. 

As AI-assisted code generation continues to influence how software is written, maintaining consistent, accurate, 
and contextually rich documentation becomes increasingly vital. The methodology and best practices outlined in 
this white paper enable teams to transform documentation from a burden into a strategic asset—fostering collaboration, 
accelerating onboarding, and ultimately enhancing the quality and longevity of software projects.


\printbibliography%
\end{document}